
\subsubsection{Disordered phase} \label{subsec:SivakDisorderedPhase}
In the disordered phase, $m_A=m_B\equiv m$ (hence $\zeta_A=\zeta_B\equiv \zeta$). 
The metric in this case reduces to
\begin{equation}
    g_{\mu\nu} \;=\; \frac{\beta\,\zeta}{\Gamma\bigl(1+\beta K \zeta\bigr)^2}
    \begin{pmatrix}
        K^2 m^2 & -K m \\[2pt]
        -K m & 1
    \end{pmatrix}.
\end{equation}
The metric is rank-one for any value of $m$, with $\det g=0$. This is physically expected: in the disorder phase the system is uniquely defined by a single order parameter -- the total magnetization, therefore there is no dissipation associate with changing the parameters such that the total magnetization stays fixed. Indeed, the null ezgenvector aligns with directions of constant $m$.
The problem of minimizing distances therefore reduces to a 1D minimization: at what distance from the phase boundary does the length minimize? 

The disordered phase there is a single nonzero eigenvalue that defines a unique geodesic submanifold. The null direction  corresponds to constant-magnetization curves, which carry zero thermodynamic cost. Hence the effective optimization reduces to minimizing the scalar prefactor of the metric in $\beta$, rather than the full Riemannian length. Geometrically, the least-dissipation trajectory is the projection of the full geodesic flow onto this one-dimensional submanifold, collapsing the dynamics of the manifold onto its non-degenerate component. In this model, the minimum occurs at $\beta=0$ (see End Matter), namely at infinite temperature. This means that the optimal trajectory between two points in this phase follows the null trajectory from the initial point to $\beta=0$, then traverse along $\beta=0$, then return along a null trajectory to the final point. 


